Eagle: multi-locus association mapping on a genome-wide scale made routine

### 2019-08-20 BIOINF-2019-0261-RevisionV2/bioinformaticspaper.tex
Morning submit
Final Revision submission - 20Aug2019

%%\usepackage{xr}
%%\renewcommand{\baselinestretch}{1.4} 
%%\usepackage{graphicx}
%%\usepackage{bm}
%%\usepackage{lineno}
%%\usepackage{natbib}
%%\usepackage{setspace}
%%\usepackage{authblk}
%%\usepackage{xcolor}

@@

%%\linenumbers

\abstract{\textbf{Motivation:} We present Eagle, a new method for multi-locus association mapping. The motivation for developing Eagle was to make multi-locus association mapping "easy" and the method-of-choice. Eagle's strengths are that it a. is considerably more powerful than single-locus association mapping b. does not suffer from multiple testing issues c. gives results that are immediately interpretable and d. has a computational footprint comparable to single-locus association mapping. \\
\textbf{Results:} By conducting a large simulation study, we will show that Eagle finds true and avoids false SNP-trait associations better than competing single- and multi-locus methods. We also analyse data from a published mouse study. Eagle found over 50\% more validated findings than the state-of-the-art single-locus method.  \\
\textbf{Availability and Implementation:} Eagle has been implemented as an R package, with a browser-based Graphical User Interface (GUI) for users less familiar with R. It is freely available via the CRAN website  at https://cran.r-project.org. Videos, Quick Start guides, FAQs, and Demos are available via the Eagle website http://eagle.r-forge.r-project.org \\
\textbf{Contact:} \href{andrew.george@csiro.au}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

\citep{Boffelli03}

@@

\subsubsection {Reducing the dimension of the model:}

@@

 \subsubsection{ Multi-locus methods:} We compare the computational and statistical performance of Eagle against five multi-locus methods. \textcolor{red}{They are bigRR  \citep{shen2013novel}, LMM-Lasso \citep{rakitsch2013lasso}, glmnet \citep{Friedman2010glmnet}, 
MLMM \citep{segura2012efficient}, and r2VIM \citep{szymczak2016r2vim}}.  All but glmnet have been purposely designed for genome-wide association mapping. 
BigRR, LMM-Lasso, and glmnet are regression-based regularisation 
methods. BigRR is based on generalised ridge regression, LMM-Lasso is based on lasso, and glmnet is based on elastic net. 
Regularisation methods make parameter estimation possible in models  where the number of predictors is far greater than the number of samples. 
They allow the strength of association between all the SNPs and trait to be measured within a single model, simultaneously. 
A limitation of these methods though is that the statistical significance of the SNP effects cannot be easily determined. 
 Due to the adaptive nature of the estimation procedures, to do this 
analytically is challenging and is an area of active research \citep{lockhart2014significance}. Instead, we calculate significance empirically via 
stability selection (see below).

@@

\subsubsection{Single-locus methods:} We also compare the performance of Eagle against two single-locus methods, GEMMA \citep{zhou2012genome} and FaST-LMM \citep{lippert2011fast}. Both are based on linear mixed models. The models have a single fixed effect for the SNP,  other fixed effects, 
a single random effect to account for familial relatedness (or polygenic background), and an error. The significance of the SNP effect in the model 
is a measure of the strength of association.  They are of the same computational complexity \citep{zhou2012genome}, and produce exact results.  
Both perform a single spectral decomposition of the relationship (or similarity) matrix $K$, use  an eigenvector matrix to rotate the data, 
and reformulate the (residual) log likelihood for easier computation. They do differ in their estimation procedure. GEMMA implements Newton-Raphson. 
FaST-LMM implements Brent's algorithm. Newton-Raphson is more complicated but has better convergence properties than Brent's algorithm. 
Both methods are state-of-the-art and have been implemented in highly efficient computer programs.

@@

\subsection{Simulation Study}
A large simulation study was performed where we sought to  answer two questions. 
First, how well does Eagle find true associations (power) and avoid 
false associations (type 1 errors)? Second, how does Eagle compare, in terms of run time and memory usage, to 
competing implementations? Data were generated under six different scenarios; a study of size 150 individuals 
and 5,000 single SNPs (150 x 5K),  350 individuals and 400,000 SNPs (350 x 400K),  1,500 individuals and 
50,000 SNPs (1500 x 50K), 2,000 individuals and 500,000 SNPs (2000 x 500K), 4,000 individuals and 
1,500,000 SNPs (4000 x 1.5M), and 10,000 individuals and 1,500,000 SNPs (10000 x 1.5M).   
These scenarios reflect, at least in some cases, the sizes of study being performed in animals, plants, and humans.

@@

\begin{figure}[!tpb]
 \centerline{\includegraphics[width=0.45\textwidth,natwidth=610,natheight=642]{Figure2_time.jpg}}
 \caption{
Median run times, in minutes, for the analysis of simulation study data from the six scenarios. Eagle is compared against five other multi-locus programs/packages (A) and two single-locus programs (B). The x- and y-axes are on a log scale for improved aesthetics. Eagle has the lowest run-times of the multi-locus programs/packages, sometimes by orders of magnitude. Eagle can even produce results faster than single-locus programs. The median run times for the programs/packages across the scenarios are given in the table (C). The entries in a blue font  correspond to the lowest run-time for a scenario. FaST-LMM$^{all}$ is where calculation of the similarity matrix is based on all the SNP data.  
 FaST-LMM$^{few}$ is where calculation  of the similarity matrix is based on a subset of the SNP data. }
\label{fig_time}
\end{figure}

@@

\textcolor{red}{
Upon submitting our paper for review, a more recent multi-locus association mapping method, FarmCPU \citep{liu2016iterative}, was brought to our attention. 
It is an unorthoﬂdox approach. Instead of working with a single model, results are passed back and forth between two models, 
a fixed effects model and a random effects model. Measures of association are obtained from the fixed effect model, which in turn 
help define pseudo-QTN from the random effects model. Conversely, pseudo-QTN found from the random effects model are passed 
back to the fixed effect model to better refine the measures of association.  The method involves multiple rounds of genome-wide 
testing. Out of interest, we reanalysed the mouse data with FarmCPU.  Runtimes were around five times longer for FarmCPU than 
Eagle.  Interestingly, FarmCPU found the same number of associations, 106,  as Eagle$^{optimal}$, but the findings were only the 
same for 11 of the traits.  For 14 traits, Eagle found more associations. For the other 14 traits, FarmCPU found more associations. 
Where the new findings from Eagle could 
be confirmed, of the 18 new associations found by FarmCPU, we were unable to confirm 10 using the results from the original study. 
Also, the two methods differ significantly in their implementation. FarmCPU is not an R package but a set of R scripts. We had 
no problem in using the scripts but this may not be true for non-R users. In contrast, Eagle has been developed for ease-of-use.
Its browser-based GUI makes it accessible to all.
}

@@

\bibitem[Bofelli {\it et~al}., 2000]{Boffelli03}
Bofelli,F., Name2, Name3 (2003) Article title, {\it Journal Name}, {\bf 199}, 133-154.

>>>>

\usepackage{lmodern}% http://ctan.org/pkg/lm
\usepackage{bm}
\usepackage{lineno}

@@

\linenumbers

\abstract{\textbf{Motivation:} We present Eagle, a new method for multi-locus association mapping. The motivation for developing Eagle was to make multi-locus association mapping "easy" and the method-of-choice. Eagle's strengths are that it a. is considerably more powerful than single-locus association mapping b. does not suffer from multiple testing issues c. gives results that are immediately interpretable and d. has a computational footprint comparable to single-locus association mapping. \\
\textbf{Results:} By conducting a large simulation study, we will show that Eagle finds true and avoids false SNP-trait associations better than competing single- and multi-locus methods. We also analyse data from a published mouse study. Eagle found over 50\% more validated findings than the state-of-the-art single-locus method.  \\
\textbf{Availability and Implementation:} Eagle has been implemented as an R package, with a browser-based Graphical User Interface (GUI) for users less familiar with R. It is freely available via the CRAN website  at https://cran.r-project.org. Videos, Quick Start guides, FAQs, and Demos are available via the Eagle website http://eagle.r-forge.r-project.org \\
\textbf{Contact:} andrew.george@csiro.au\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

@@

\subsubsection {Reducing the dimension of the model}

@@

 \subsubsection{ Multi-locus methods}

 We compare the computational and statistical performance of Eagle against five multi-locus methods. \textcolor{red}{They are bigRR  \citep{shen2013novel}, LMM-Lasso \citep{rakitsch2013lasso}, glmnet \citep{Friedman2010glmnet}, 
MLMM \citep{segura2012efficient}, and r2VIM \citep{szymczak2016r2vim}}.  All but glmnet have been purposely designed for genome-wide association mapping. 
BigRR, LMM-Lasso, and glmnet are regression-based regularisation 
methods. BigRR is based on generalised ridge regression, \hbox{LMM-Lasso} is based on lasso, and glmnet is based on elastic net. 
Regularisation methods make parameter estimation possible in models  where the number of predictors is far greater than the number of samples. 
They allow the strength of association between all the SNPs and trait to be measured within a single model, simultaneously. 
A limitation of these methods though is that the statistical significance of the SNP effects cannot be easily determined. 
 Due to the adaptive nature of the estimation procedures, to do this 
analytically is challenging and is an area of active research \citep{lockhart2014significance}. Instead, we calculate significance empirically via 
stability selection (see below).

@@

\subsubsection{Single-locus methods} 
We also compare the performance of Eagle against two single-locus methods, GEMMA \citep{zhou2012genome} and FaST-LMM \citep{lippert2011fast}. Both are based on linear mixed models. The models have a single fixed effect for the SNP,  other fixed effects, 
a single random effect to account for familial relatedness (or polygenic background), and an error. The significance of the SNP effect in the model 
is a measure of the strength of association.  They are of the same computational complexity \citep{zhou2012genome}, and produce exact results.  
Both perform a single spectral decomposition of the relationship (or similarity) matrix $K$, use  an eigenvector matrix to rotate the data, 
and reformulate the (residual) log likelihood for easier computation. They do differ in their estimation procedure. GEMMA implements Newton-Raphson. 
FaST-LMM implements Brent's algorithm. Newton-Raphson is more complicated but has better convergence properties than Brent's algorithm. 
Both methods are state-of-the-art and have been implemented in highly efficient computer programs.

@@

\subsection{Simulation Study}

A large simulation study was performed where we sought to  answer two questions. 
First, how well does Eagle find true associations (power) and avoid 
false associations (type 1 errors)? Second, how does Eagle compare, in terms of run time and memory usage, to 
competing implementations? Data were generated under six different scenarios; a study of size 150 individuals 
and 5,000 single SNPs (150 x 5K),  350 individuals and 400,000 SNPs (350 x 400K),  1,500 individuals and 
50,000 SNPs (1500 x 50K), 2,000 individuals and 500,000 SNPs (2000 x 500K), 4,000 individuals and 
1,500,000 SNPs (4000 x 1.5M), and 10,000 individuals and 1,500,000 SNPs (10000 x 1.5M).   
These scenarios reflect, at least in some cases, the sizes of study being performed in animals, plants, and humans.

@@

\begin{figure}
 \centerline{\includegraphics[width=0.45\textwidth,natwidth=610,natheight=642]{Figure2_time.jpg}}
 \caption{
Median run times, in minutes, for the analysis of simulation study data from the six scenarios. Eagle is compared against five other multi-locus programs/packages (A) and two single-locus programs (B). The x- and y-axes are on a log scale for improved aesthetics. Eagle has the lowest run-times of the multi-locus programs/packages, sometimes by orders of magnitude. Eagle can even produce results faster than single-locus programs. The median run times for the programs/packages across the scenarios are given in the table (C). The entries in a blue font  correspond to the lowest run-time for a scenario. FaST-LMM$^{all}$ is where calculation of the similarity matrix is based on all the SNP data.  
 FaST-LMM$^{few}$ is where calculation  of the similarity matrix is based on a subset of the SNP data. }
\label{fig_time}
\end{figure}

@@

\textcolor{red}{
Upon submitting our paper for review, a more recent multi-locus association mapping method, FarmCPU \citep{liu2016iterative}, was brought to our attention. 
It is a statistically unorthodox approach. Instead of working with a single model, results are passed back and forth between two models, 
a fixed effects model and a random effects model. Measures of association are obtained from the fixed effect model, which in turn 
help define pseudo-QTN from the random effects model. Conversely, pseudo-QTN found from the random effects model are passed 
back to the fixed effect model to better refine the measures of association.  The method involves multiple rounds of genome-wide 
testing. Out of interest, we reanalysed the mouse data with FarmCPU.  Runtimes were around five times longer for FarmCPU than 
Eagle.  Interestingly, FarmCPU found the same number of associations, 106,  as Eagle$^{optimal}$, but the findings were only the 
same for 11 of the traits.  For 14 traits, Eagle found more associations. For the other 14 traits, FarmCPU found more associations. 
Where the new findings from Eagle could 
be confirmed, of the 18 new associations found by FarmCPU, we were unable to confirm 10 using the results from the original study. 
Also, the two methods differ significantly in their implementation. FarmCPU is not an R package but a set of R scripts. We had 
no problem in using the scripts but this may not be true for non-R users. In contrast, Eagle has been developed for ease-of-use.
Its browser-based GUI makes it accessible to all.
}

@@
